Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques

### 2019-06-11 paper/paperV1.tex
ML Corrections

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\begin{document}
\author{F\'elix Gontier $^1$, Catherine Lavandier $^2$, Pierre Aumond $^{2, 3}$\\Mathieu Lagrange $^1$, Jean-Francois Petiot $^1$}
\date{
$^1$ LS2N, UMR 6004 CNRS, Ecole Centrale de Nantes, F-44321, France\\
$^2$ ETIS, UMR 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
$^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
}
\title{Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques}
\maketitle

@@

In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}. Large amounts of data with task-specific annotations are needed for these architectures to learn to extract relevant information from recordings. To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with perceptual activity labels. We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.

@@

In order to gather estimates of the perceputal attributes and to validate the usefulness of the proposed corpus, a perceptual experiment is conducted. Participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

@@

The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used for all participants. These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listeners.

@@

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_rec} as red dots for simulated scenes corpus and blue dots for recorded and replicated scenes. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.

@@

\begin{figure}[th]
    \centering
    \includegraphics[width=\columnwidth]{figures/pca_p1.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25). Red dots are projections of the assessments of each scene averaged over subjects. Arrows show discrepancies between corresponding recorded and replicated scenes. For the P1 recording location the standard deviation of the projection of individual responses is shown using ellipses.\ml{Pas de couleur, Remplacer par des croix et des points !! enlever le blanc autour de la figure, soit a la main, soit en controlant la bounding box}}\label{fig:pspace_rec}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=\columnwidth]{figures/pca_sim.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75). Assessments of simulated scenes are projected as red dots, and recorded and replicated scenes as blue dots.}\label{fig:pspace_sim}
\end{figure}

@@

The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity. Its expression is:
\begin{equation}
TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f1=31.5Hz}^{f1=16kHz}\lvert\frac{d^2L(f, t)}{dfdt}\rvert}
\end{equation}
where $L(f, t)$ is the third-octave spectrum of the signal. To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) as an estimate the contribution of birds. The 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) is chosen as an estimate the contribution of voices.\\

@@

Target outputs are associated to each input frame to train the model. In~\cite{gontier2018} the time of presence of traffic, voices and birds was successfully approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values given by this indicator can be used as weak perceptual time of presence labels. Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha_{opt}, \beta_{opt})$. % Though, training a deep learning model for sound level regression is more difficult: it requires the architecture to perform implicit source separation, whereas presence detection does not need this information to be extracted from the data. This will not be discussed here as it did not result in better time of presence estimations in our experiments.

A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}. The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$. The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. A final sigmoid activation is used in order to obtain outputs in the 0-1 range. The resulting 3 values correspond to the presence of traffic, voices and birds in the 1~s frame respectively. During training these values are directly compared to presence labels given by $\hat T_s(\alpha_{opt}, \beta_{opt})$ using a binary cross-entropy cost function:
\begin{equation}
BCE(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{equation}
where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range. This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples. During evaluation, a threshold of 0.5 is independently applied to the 3 outputs to obtain a binary presence value for each source: each source is considered absent when the model outputs a value lower than 0.5 and present when it outputs 0.5 or higher. The time of presence estimation is then obtained by averaging presence labels of all 1~s time frames corresponding to the same scene.

\begin{figure}[t]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/deep_arch.pdf}
    \caption{Architecture of the deep learning model used for source presence prediction.\ml{refaire la figure en vertical}}\label{fig:deep_arch}
\end{figure}

@@

\begin{table*}
\centering
\caption{Pearson's correlation coefficients between physical and perceptual indicators (n = 92, *: p<0.05, **: p<0.01). Non significant correlations at the 5\% threshold are noted NS.}
\label{tab:physc}
\resizebox{\textwidth}{!}{
\begin{tabular}{ c | c c c c c | c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	$LA_{eq}$ & -0.86** & 0.68** & 0.92** & -0.37** & -0.88** & 0.77** & 0.66** & NS & -0.41** \\
	$LA_{50}$ & -0.84** & 0.67** & 0.91** & -0.33** & -0.87** & 0.71** & 0.63** & NS & -0.35** \\
	$L_{eq}$ & -0.88** & 0.67** & 0.91** & -0.44** & -0.88** & 0.83** & 0.71** & NS & -0.46** \\
	$L_{10}$ & -0.87** & 0.65** & 0.90** & -0.44** & -0.86** & 0.84** & 0.71** & NS & -0.47** \\
	$L_{50}$ & -0.89** & 0.65** & 0.92** & -0.43** & -0.89** & 0.77** & 0.71** & NS & -0.44** \\
	$L_{90}$ & -0.86** & 0.68** & 0.92** & -0.39** & -0.89** & 0.71** & 0.67** & NS & -0.40** \\
	$L_{50, 1kHz}$ & -0.88** & 0.69** & 0.92** & -0.42** & -0.89** & 0.74** & 0.73** & NS & -0.50** \\
	$L_{10}-L_{90}$ & NS & NS & -0.24* & NS & -0.22* & NS & NS & NS & NS \\ \hline
	$TFSD_{500Hz, 1s}$ & NS & 0.41** & NS & 0.28** & NS & -0.24* & -0.39** & 0.74** & NS \\
	$TFSD_{4kHz, 1/8s}$ & 0.49** & -0.41** & -0.45** & 0.35** & 0.50** & -0.41** & -0.48** & -0.21* & 0.61** \\ \hline
	$L_{eq, T}$ & -0.58** & NS & 0.46** & -0.46** & -0.42** & 0.77** & 0.71** & NS & -0.36** \\
	$L_{eq, V}$ & NS & 0.50** & 0.31** & NS & -0.37** & NS & NS & 0.71** & -0.40** \\
	$L_{eq, B}$ & 0.27* & NS & NS & 0.35** & NS & -0.25* & -0.24* & NS & 0.71** \\ \hline
	$\Delta L_T$ & -0.45** & NS & 0.26* & -0.59** & -0.22* & 0.63** & 0.66** & -0.51** & -0.26* \\
	$\Delta L_V$ & NS & 0.50** & NS & 0.35** & NS & -0.27** & -0.38** & 0.59** & NS \\
	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & NS & 0.25* & -0.24* & -0.25* & NS & NS \\ \hline
	$\hat T_T(\alpha_{opt}, \beta_{opt})$ & -0.51** & NS & 0.33** & -0.56** & -0.27** & 0.58** & 0.82** & -0.36** & -0.33** \\
	$\hat T_V(\alpha_{opt}, \beta_{opt})$ & NS & 0.44** & NS & 0.36** & NS & -0.23* & -0.41** & 0.83** & NS \\
	$\hat T_B(\alpha_{opt}, \beta_{opt})$ & 0.58** & -0.33** & -0.47** & 0.55** & 0.52** & -0.45** & -0.56** & NS & 0.91** \\ \hline
\end{tabular}
}
\end{table*}

Subsequently, multilinear models of pleasantness are constructed based on the results of the analysis of correlations. As in Section~\ref{sec:results_perc}, all combinations of the proposed physical variables are considered, and a VIF check is performed on predictors to ensure that no multi-collinearity between predictors exists in a model. On the present corpus, the best model in terms of $R^2_{adj}$ is:
\begin{equation}
\hat P_{1, \varphi} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
\end{equation}
where $\varphi$ indicates a model from physical variables. Indicators related to traffic and voices activity are absent compared to the perceptual models in Section~\ref{sec:results_perc}. This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{1, p}$ (eq.\ref{eq:pp1}) and $\hat P_{2, p}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.

This model is compared to two baselines proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, for which predictor variables are directly computed from the audio signal. Both models' coefficients are re-optimized on the studied data for a fair comparison:
\begin{align}
\hat P_{2, \varphi} &= 18.67 - 0.20~L_{50} - 0.02~(L_{10}-L_{90})\\
\hat P_{3, \varphi} &= 13.03 - 0.16~L_{50, 1kHz} + 9.60~TFSD_{500Hz, 1s} \\
& + 5.00~TFSD_{4kHz, 1/8s}
\end{align}

@@

\begin{table}[t]
\centering
\caption{Performance of physical models for pleasantness prediction (**: p<0.01).}
\label{tab:physm}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c | c | c }
\hline
	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
	$\hat P_{1, \varphi}$ & 0.83 & 0.82 & 0.91** \\
	$\hat P_{2, \varphi}$ & 0.90 & 0.79 & 0.89** \\
	$\hat P_{3, \varphi}$ & 0.90 & 0.78 & 0.89** \\ \hline
\end{tabular}
%}
\end{table}

@@

\begin{table*}[t]
\centering
\caption{Pleasantness prediction quality on the perceptual experiment corpus using the source detection model compared to labels. The corpus is split in three parts: the 6 recorded scenes (Rec.), the 19 replicated scenes (Rep.), and the 75 scenes with simulated scenarios (Sim.).}
\label{tab:pppred}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c | c c c }
\hline
	Model & \multicolumn{4}{|c}{$P_{1, \varphi}$ with model outputs} & \multicolumn{3}{|c}{$P_{1, \varphi}$ with $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels} \\ \hline
	Sub-corpus & All & Rec. & Rep. & Sim. & All & Rep. & Sim. \\ \hline
	RMSE & 0.84 & 1.09 & 0.68 & 0.85 & 0.83 & 0.73 & 0.85 \\ \hline
	r & 0.91** & 0.89** & 0.93** & 0.89** & 0.91** & 0.92** & 0.89** \\ \hline
	$R^2_{adj}$ & 0.82 & 0.73 & 0.82 & 0.80 & 0.82 & 0.81 & 0.79 \\ \hline
\end{tabular}
%}
\end{table*}

@@

The perceptual models of pleasantness (eq.\ref{eq:pp1}-\ref{eq:pp2}) are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):
\begin{align}
\hat P_{3, p} &= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} \\
& + 0.15~T_{B, p}\\
\hat P_{4, p} &= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} \\
& + 0.09~T_{B, p}
\end{align}
Several hypotheses can be drawn from this result. First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the \textit{ex situ} setting of the experiment. However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for \textit{in situ} models. This may be due to scenes at the same sound level appearing louder when played through headphones in a laboratory setting.

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\begin{document}
\author{F\'elix Gontier $^1$, Catherine Lavandier $^2$, Pierre Aumond $^{2, 3}$\\Mathieu Lagrange $^1$, Jean-Francois Petiot $^1$}
\date{
$^1$ LS2N, UMR 6004 CNRS, Ecole Centrale de Nantes, F-44321, France\\
$^2$ ETIS, UMR 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
$^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
}
\title{Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques}
\maketitle

@@

In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}. Large amounts of data with task-specific annotations are needed for these architectures to learn to extract relevant information from recordings. To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with perceptual activity labels. We propose in this paper to study the potential of the use of simulated corpora~\cite{lafay2016, salamon2017} for such purpose.

@@

In order to gather estimates of the perceptual attributes and to validate the usefulness of the proposed corpus, a perceptual experiment is conducted. Participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

@@

The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used for all participants. These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener.

@@

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_sim} as dots for simulated scenes corpus and crosses for recorded and replicated scenes. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.

@@

\begin{figure}[th]
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/pca_p1.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25). Arrows show discrepancies between corresponding recorded and replicated scenes. For the P1 recording location the standard deviation of the projection of individual responses is shown using ellipses.}\label{fig:pspace_rec}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/pca_sim.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75). Assessments of simulated scenes are projected as dots, and recorded and replicated scenes are projected as crosses.}\label{fig:pspace_sim}
\end{figure}

@@

The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity. Its expression is:

\begin{align}
TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f1=31.5Hz}^{f1=16kHz}\lvert\frac{d^2L(f, t)}{dfdt}\rvert}
\end{align}

where $L(f, t)$ is the third-octave spectrum of the signal. To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) as an estimate the contribution of birds. The 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) is chosen as an estimate the contribution of voices.\\

@@

Target outputs are associated to each input frame to train the model. In~\cite{gontier2018} the time of presence of traffic, voices and birds was successfully approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values given by this indicator can be used as weak perceptual time of presence labels. Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha_{opt}, \beta_{opt})$.

A deep learning model is implemented and trained for the source presence prediction task, with the architecture shown in Figure~\ref{fig:deep_arch}. The model includes 4 blocks of convolutional layers followed by leaky rectified linear unit (LeakyReLU) activations of expression $y = max(0.1x, x)$. The convolutional layers have respectively 128, 64, 32 and 32 output maps, \textit{i.e.} channels, and a common kernel size of 5x5. The output of the last block is flattened then goes through a fully connected layer with output size 3. A final sigmoid activation is used in order to obtain outputs in the 0-1 range. The resulting 3 values correspond to the presence of traffic, voices and birds in the 1~s frame respectively. During training these values are directly compared to presence labels given by $\hat T_s(\alpha_{opt}, \beta_{opt})$ using a binary cross-entropy cost function:

\begin{align}
BCE(y, \hat y) = -\sum_s y_s log\left(\hat y_s\right) + (1-y_s) log\left(1-\hat y_s\right)
\end{align}

where $s$ is the source, $y_s$ and $\hat y_s$ are the target and predicted presence for source $s$ in the 0-1 range. This loss function is minimized using the Adam algorithm~\cite{kingma2015} on batches of 1~s examples. During evaluation, a threshold of 0.5 is independently applied to the 3 outputs to obtain a binary presence value for each source: each source is considered absent when the model outputs a value lower than 0.5 and present when it outputs 0.5 or higher. The time of presence estimation is then obtained by averaging presence labels of all 1~s time frames corresponding to the same scene.

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figures/deep_arch.pdf}
    \caption{Architecture of the deep learning model used for source presence prediction.}\label{fig:deep_arch}
\end{figure}

@@

\begin{table*}
\centering
\caption{Pearson's correlation coefficients between physical and perceptual indicators (n = 92, *: p<0.05, **: p<0.01). Non significant correlations at the 5\% threshold are noted NS.}
\label{tab:physc}
\resizebox{\textwidth}{!}{
\begin{tabular}{ c | c c c c c | c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	$LA_{eq}$ & -0.86** & 0.68** & 0.92** & -0.37** & -0.88** & 0.77** & 0.66** & NS & -0.41** \\
	$LA_{50}$ & -0.84** & 0.67** & 0.91** & -0.33** & -0.87** & 0.71** & 0.63** & NS & -0.35** \\
	$L_{eq}$ & -0.88** & 0.67** & 0.91** & -0.44** & -0.88** & 0.83** & 0.71** & NS & -0.46** \\
	$L_{10}$ & -0.87** & 0.65** & 0.90** & -0.44** & -0.86** & 0.84** & 0.71** & NS & -0.47** \\
	$L_{50}$ & -0.89** & 0.65** & 0.92** & -0.43** & -0.89** & 0.77** & 0.71** & NS & -0.44** \\
	$L_{90}$ & -0.86** & 0.68** & 0.92** & -0.39** & -0.89** & 0.71** & 0.67** & NS & -0.40** \\
	$L_{50, 1kHz}$ & -0.88** & 0.69** & 0.92** & -0.42** & -0.89** & 0.74** & 0.73** & NS & -0.50** \\
	$L_{10}-L_{90}$ & NS & NS & -0.24* & NS & -0.22* & NS & NS & NS & NS \\ \hline
	$TFSD_{500Hz, 1s}$ & NS & 0.41** & NS & 0.28** & NS & -0.24* & -0.39** & 0.74** & NS \\
	$TFSD_{4kHz, 1/8s}$ & 0.52** & -0.43** & -0.49** & 0.41** & 0.52** & -0.45** & -0.54** & NS & 0.63** \\ \hline
	$L_{eq, T}$ & -0.58** & NS & 0.46** & -0.46** & -0.42** & 0.77** & 0.71** & NS & -0.36** \\
	$L_{eq, V}$ & NS & 0.50** & 0.31** & NS & -0.37** & NS & NS & 0.71** & -0.40** \\
	$L_{eq, B}$ & 0.27* & NS & NS & 0.35** & NS & -0.25* & -0.24* & NS & 0.71** \\ \hline
	$\Delta L_T$ & -0.45** & NS & 0.26* & -0.59** & -0.22* & 0.63** & 0.66** & -0.51** & -0.26* \\
	$\Delta L_V$ & NS & 0.50** & NS & 0.35** & NS & -0.27** & -0.38** & 0.59** & NS \\
	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & NS & 0.25* & -0.24* & -0.25* & NS & NS \\ \hline
	$\hat T_T(\alpha_{opt}, \beta_{opt})$ & -0.53** & NS & 0.35** & -0.57** & -0.29** & 0.56** & 0.81** & -0.39** & -0.37** \\
	$\hat T_V(\alpha_{opt}, \beta_{opt})$ & NS & 0.44** & NS & 0.35** & NS & NS & -0.39** & 0.81** & NS \\
	$\hat T_B(\alpha_{opt}, \beta_{opt})$ & 0.56** & -0.30** & -0.46** & 0.55** & 0.51** & -0.46** & -0.57** & NS & 0.91** \\ \hline
\end{tabular}
}
\end{table*}

Subsequently, multilinear models of pleasantness are constructed based on the results of the analysis of correlations. As in Section~\ref{sec:results_perc}, all combinations of the proposed physical variables are considered, and a VIF check is performed on predictors to ensure that no multi-collinearity between predictors exists in a model. On the present corpus, the best model in terms of $R^2_{adj}$ is:

\begin{align}
\hat P_{1, \varphi} = 16.74 - 0.18~L_{50} + 1.01~\hat T_B(\alpha_{opt}, \beta_{opt}) \label{eq:pphi}
\end{align}

where $\varphi$ indicates a model from physical variables. Indicators related to traffic and voices activity are absent compared to the perceptual models in Section~\ref{sec:results_perc}. This is because the $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{1, p}$ (eq.\ref{eq:pp1}) and $\hat P_{2, p}$ (eq.\ref{eq:pp2}) respectively, and no strong contribution of objective variables representative of voices to pleasantness is identified in this corpus.

This model is compared to two baselines proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, for which predictor variables are directly computed from the audio signal. Both models' coefficients are re-optimized on the studied data for a fair comparison:

\begin{align}
\hat P_{2, \varphi} &= 18.67 - 0.20~L_{50} - 0.02~(L_{10}-L_{90})\\
\begin{split}
\hat P_{3, \varphi} &= 30.18 - 0.16~L_{50, 1kHz} + 8.92~TFSD_{500Hz, 1s} \\
& + 2.99~TFSD_{4kHz, 1/8s}
\end{split}
\end{align}

@@

\begin{table}[t]
\centering
\caption{Performance of physical models for pleasantness prediction (**: p<0.01).}
\label{tab:physm}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c | c | c }
\hline
	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
	$\hat P_{1, \varphi}$ & 0.83 & 0.82 & 0.91** \\
	$\hat P_{2, \varphi}$ & 0.90 & 0.79 & 0.89** \\
	$\hat P_{3, \varphi}$ & 0.91 & 0.78 & 0.89** \\ \hline
\end{tabular}
%}
\end{table}

@@

\begin{table*}[t]
\centering
\caption{Pleasantness prediction quality on the perceptual experiment corpus using the source detection model compared to labels. The corpus is split in three parts: the 6 recorded scenes (Rec.), the 19 replicated scenes (Rep.), and the 75 scenes with simulated scenarios (Sim.).}
\label{tab:pppred}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c | c c c }
\hline
	Model & \multicolumn{4}{|c}{$P_{1, \varphi}$ with model outputs} & \multicolumn{3}{|c}{$P_{1, \varphi}$ with $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels} \\ \hline
	Sub-corpus & All & Rec. & Rep. & Sim. & All & Rep. & Sim. \\ \hline
	RMSE & 0.84 & 1.09 & 0.68 & 0.85 & 0.83 & 0.72 & 0.86 \\ \hline
	r & 0.91** & 0.89** & 0.93** & 0.89** & 0.91** & 0.92** & 0.89** \\ \hline
	$R^2_{adj}$ & 0.82 & 0.73 & 0.82 & 0.80 & 0.82 & 0.79 & 0.79 \\ \hline
\end{tabular}
%}
\end{table*}

@@

The perceptual models of pleasantness (eq.\ref{eq:pp1}-\ref{eq:pp2}) are comparable to those proposed in~\cite{ricciardi2014} and~\cite{aumond2017}, of respective expressions (translated from 1-11 scales to 0-10):

\begin{align}
\begin{split}
\hat P_{3, p} {}&= 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} \\
& + 0.15~T_{B, p}
\end{split}\\
\begin{split}
\hat P_{4, p} {}&= 8.70 - 0.47~OL - 0.21~T_{T, p} + 0.12~T_{V, p} \\
& + 0.09~T_{B, p}
\end{split}
\end{align}

Several hypotheses can be drawn from this result. First, the use of a simulated corpus with low source complexity does not alter the fundamental relationships between perceptual quantities, nor does the \textit{ex situ} setting of the experiment. However, the latter may have an impact on the contribution of the overall loudness to pleasantness: the regression coefficient is $-0.67$, which is significantly higher than for \textit{in situ} models. This may be due to scenes at the same sound level appearing louder when played through headphones in a laboratory setting.
